A* with perfect potentials

### 2019-10-23 paper/ch_potentials.tex
add missing comma
write down stuff for arxiv version

\title{Dummy title} %TODO Please add

\titlerunning{Dummy short title}%optional, please use if title is longer than one line

@@

\authorrunning{B. Strasser and T.\, Zeitz}%TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

@@

%TODO mandatory: add short abstract of the document
\begin{abstract}
Lorem ipsum dolor sit amet, consectetur adipiscing elit. Praesent convallis orci arcu, eu mollis dolor. Aliquam eleifend suscipit lacinia. Maecenas quam mi, porta ut lacinia sed, convallis ac dui. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Suspendisse potenti.
\end{abstract}

@@

>>>>

\title{A* with perfect potentials} %TODO Please add

% \titlerunning{Dummy short title}%optional, please use if title is longer than one line

@@

\authorrunning{B. Strasser and T. Zeitz}%TODO mandatory. First: Use abbreviated first/middle names. Second (only in severe cases): Use first author plus 'et al.'

@@

\begin{abstract}
A* is an extension to Dijkstra's algorithm.
It uses a potential function to estimate the distance of each node to the target.
By adding these estimates to the queue keys, the search is directed towards the target.
The quality of this potential determines the performance of A*.
We introduce a novel potential function which utilizes Contraction Hierarchies to efficiently calculate perfect potentials on road networks.
Perfect potentials can be used to speed up shortest path queries for extended problem settings like for example frequently changing or time-dependent edge weights.
\end{abstract}

@@

Fast routing in huge street networks has seen a lot of research \cite{Survey}.
For static, time- and user-invariant networks a lot of techniques exist.
They work in two phases.
The first phase is called preprocessing phase and is a slow index-construction.
In this phase the road graph and the edge weights are known.
In the second phase, called query phase, minimum weight paths are computed.
The query phase has access to the precomputed index and is very fast.

One of the most successful ones are Contraction Hierarchies~(CH)~\cite{CH}.
Contraction Hierarchies exploit inherent hierarchies in the shortest-path structure of road networks.
Using CHs, Queries can be answered on the order of only a millisecond.
However, CHs can not always be easily extended to more complicated problem settings.
% TODO what problem settings
While such extensions exist~\cite{CCH,KaTCH}, they often require substantial engineering effort.

On the other hand, goal directed speed-up techniques based on A* are usually slower but more flexible.
The performance of these approaches depends on the quality of the potential functions used for the A* search.
In this paper, we propose an algorithm which utilizes Contraction Hierarchies to efficiently calculate perfect (best possible) potentials.
This allows us to retain the flexibility of A* based approaches but use much more effective potentials than previously possible.
% These techniques are used in online navigation services.
% In these services, there is a single index.
% With this index a large number of user requests are answered.
% The preprocessing phases is rerun on a regular interval to incorporate map updates.

% While highly scalable, this setup has the disadvantage, that every user must have use the same weights.
% In this paper, we explore an novel algorithm, called CH-Potentials, that uses the two phase setup, however, allows the weights to be modified on a per query basis.
% Applications for this algorithm are personalized user settings such as for example:
% Avoiding highways, capping the maximum speed, or avoid user-configurable areas.

\paragraph*{Related Work}

% We are not the first to investigate this problem setting.
% Two approaches exist in the literature.

% The three phase approach splits the preprocessing phase into two.
% Techniques that use this setup are MLD/CRP \cite{?} and CCH \cite{?}.
% The first phase is a weight-independent preprocessing phase that analsyse the graph topology.
% The second phase is called customization phase and introduces the weights into the index.
% The third phase is the query phase that computes paths.
% The preprocessing phase is very slow.
% The customization phase runs on the order of seconds.
% The query phase runs within milliseconds.

% The three phase approach is ideal to incorporate weight changes that apply to all users.
% Live traffic information is a prime example.
% It has been suggested to use the three phase approach to handle user-specific requirements by storing a copy of the index per user.
% The problem with this approach is that it binds a lot of server-side resources per user.
% Further, if the user often changes his weights, the customization needs to run often.
% Every customization requires seconds of running time.
% Just applying Dijkstra's algorithm to the input graph results in similar running times.

% The second approach is Sabine's VLDB personalized routing.
% Also TopoCore~\cite{?}.

A plethora of of research exists in the area of efficient routing in road networks \cite{Survey}.
Here, we are going to focus on the algorithms relevant to this work.

At the core is the algorithm of Dijkstra~\cite{Dijkstra}.
This algorithm explores the graph starting at a source node $s$ and visiting nodes ordered by increasing distance from $s$.
For each node $v$, a tentative distance $d_s(v)$ is maintained, which is initially set to $\infty$.
In each iteration the node $u$ with the smallest remaining distance $d_s(u)$ is extracted from a priority queue and settled.
Each outgoing edge $(u,v)$ of $u$ is relaxed, that is to check if $d_s(u) + w(u,v) < d_s(v)$ and if so update the distance and the queue position of $v$ or insert $v$ into the queue.
Once the target node $t$ is settled, the shortest path and distance between $s$ and $t$ is known.

A*~\cite{?} extends Dijkstra's algorithm by changing the order in which nodes are visited such that nodes in the direction of $t$ are visited earlier.
This is achieved through a potential function $p$ which estimates for each node the distance to $t$.
The priority queue is then ordered by $d_s(u) + p(u)$.

ALT~\cite{ALT} is an A* based technique which precomputes shortest distances between a small set of landmark nodes and all other nodes during preprocessing.
For the query, potentials are calculated using the triangle inequality for shortest distances and the precomputed distances.
This achieves decent speed-ups and can be used in other problem settings but is not competitive to hierarchical approaches

Contraction Hierarchies~\cite{CH} are a popular hierarchical speed-up technique.
During preprocessing, all nodes are ranked by some measure of importance.
Nodes which lie on many shortest paths (for example highway nodes) are important.
Then, nodes are contracted iteratively by increasing rank.
To contract a node $v$ means removing it form the graph, but preserving all shortest distances among all other nodes.
This is achieved by inserting shortcut edges between the neighbors of $v$ which have the length of the path over $v$, if necessary.
The query is a bidirectional Dijkstra starting from $s$ and $t$ where only edges to nodes with higher rank are relaxed.

PHAST~\cite{?} is an algorithm for one-to-many queries based on Contraction Hierarchies.
The preprocessing phase remains the same.
For the query, the forward search $s$ also remains the same.
Instead of running a backward upward search from all nodes, PHAST now relaxes all edges $(u,v)$ in the backward graph where $rank(u) > rank(v)$ ordered descending by $rank(v)$.
To achieve good running times, it is important that nodes and edges are ordered such that relaxing the downward edges is just a linear sweep over the edge array.

% \section{Preliminaries}
% \label{sec:prelim}

\section{Perfect Potentials}
\label{sec:algo}

Our new algorithm is at its core a combination of A* and PHAST.
% For an additional performance, we further combine it with TopoCore.

The input of the preprocessing phase is a graph and lower-bound weights $w_{\ell}$.
Denote by $d_{\ell}(x,y)$ the minimum weight $xy$-path distance with respect to $w_{\ell}$.
The preprocessing phase consists of computing a CH of the graph with lower-bound weights.

The input to the query phase is a source $s$ and a target $t$ node and weights $w_u$.
These weights $w_u$ must not be smaller than the lower-bound weights $w_{\ell}$, i.e., $w_{\ell}(e) < w_u(e)$ for all edges $e$.
The weights may be time-dependent travel time predictions, incorporate live traffic or user specific preferences.
Analogous to $d_{\ell}(x,y)$, $d_u(x,y)$ is the minimum-weight phase according to $w_u$.
We obtain $d_u(x,y)$ through an A* search with respect to $d_u$ from $s$ to $t$.
The distance $d_{\ell}(x,t)$ from a node $x$ to $t$ is used as potential.

\begin{algorithm2e}
\KwData{$B[x]$ distance from $s$ to $x$ compute in backward search, or $+\infty$ if unreachable}
\KwData{$P[x]$ memoized potential for $x$, or $\bot$ if not yet computed. Equal to $d_{\ell}(x,t)$}
\SetKwFunction{Pot}{Potential}
\SetKwProg{Fn}{Function}{:}{}
\Fn{\Pot{$x$}}{
  \If{$P[x] = \bot$}{
    $P[x]\leftarrow B[x]$\;
    \For{$(x,y)$ is upward CH edge with weight $w_{xy}$}{
      $P[x]\leftarrow \min\{P[x],w_{xy}+\Pot(y)\}$\;
    }
  }
  \Return{$P[x]$}\;
}

\caption{Algorithm to compute Potentials.}
\label{algo:pot}

\end{algorithm2e}

The core contribution of our algorithm consist in the way $d_{\ell}(x,t)$ is efficiently computed.
For this, we use a PHAST variant.
In the first step, we explore all nodes backward-reachable from $t$ in the CH.
We store the computed distances in the array $B$.
We then execute the A* search.
Potentials are computed recursively, on-the-fly, with memoization as depicted in Algorithm \ref{algo:pot}.
If a potential is unknown, it is computed by recursively computing the potentials of all higher nodes in the CH backward search graph.
These potentials are then increased by the distance in the CH backwards search graph.
The potential of a node $x$ is the minimum over all these values and the distance found in the CH forward search.

% Our algorithm can be applied as described to the input graph.
% However, depending on the weights, the A* search can get stuck often in deadends.
% We therefore first coarsen the input graph using TopoCore and apply the above described algorithm to the core.
% In this setup, TopoCore handles all deadends.

% \section{Applications}
% \label{sec:applications}

% \subsection{Live Traffic}
% \subsection{Time-dependent Routing}

% \section{Experiments}
% \label{sec:exp}

\section{Conclusion}
\label{sec:conclusion}

We introduce a novel way to compute potentials for the A* algorithm.
The potentials are calculated utilizing a CH on a lower bound graph and perfect with regards to that lower bound graph.
With these potentials we can speed up A* search for extended route planning problems like route planning with live traffic, time-dependent travel time predictions or user preferences.
